Future challenges for Research Software Engineering

### 2025-09-23 correspondence/rebuttal-1.md

We reworded it as a statement.

> As per previous comments, wording like “we expect/hope” should be removed from the Achieved Goals box

Thanks for  spotting this. We increased our boldness and removed every hope...

> Page 10
> Life would be so much easier if at least university regulations would be homogenised, no? - reword as Life would be so much easier if at least university regulations were homogenised.

fixed as per comment.

> As per previous comments, wording like “we would like to see” should be removed from the Achieved Goals box

Reworded the box accordingly.

> Page 11
> Nevertheless, there are some things that we declared out of scope here. Among that is the ongoing digital transformation in wider society. – Reword as one sentence

Reworded the sentence accordingly.

> Page 12
> In summary, RSEs will very likely be doing different work
> to what they are doing today - remove this line break

Not sure, how this crept into the final version... It's gone now.

### 2025-09-23 future_challenges_for_RSEng.tex

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
\texorpdfstring{\\}{} % line break
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
\texorpdfstring{\\}{} % line break
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10}\textsuperscript{,}\authorRef{11},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{12}\textsuperscript{,}\authorRef{11},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} de-RSE e.V.---Society for Research Software in Germany\\
\autlabel{12} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
\autlabel{13} Independent researcher, Germany\\
\autlabel{14} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and the research software community is, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

Further risks that were later discussed include an intensified replication crisis,
a de-democratization of research, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
and LLM-assisted data analysis to become less reproducible.
This is not an entirely new issue, and similar concerns exist in the HPC community,
where reproducing simulation data and data analysis workflows presents tough challenges~\cite{Antunes2024}.
Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
When applicable, standardized and version-controlled models should be preferred
over self-developed models, and self-developed models should be treated like any
other research output (i.e. publish the trained model, training data and training workflow).
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more likely to be driven by institutions that are able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}
and the malicious use of hidden prompts~\cite{Gibney2025}.
As researchers increase their reliance on AI assistants,
RSEs will be freed from repeatedly crafting similar individual elements,
but will be spending more of their time designing and reviewing software written by other humans and AI tools.
To address these risks, the RSE of the future should not only steer every part
of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
but also contribute to endangered community resources and infrastructure that
will keep research accessible to the wider scientific community.

@@

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}

@@

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
\texorpdfstring{\\}{} % line break
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
\texorpdfstring{\\}{} % line break
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10}\textsuperscript{,}\authorRef{11},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{12}\textsuperscript{,}\authorRef{11},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} de-RSE e.V.---Society for Research Software in Germany\\
\autlabel{12} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
\autlabel{13} Independent researcher, Germany\\
\autlabel{14} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and the research software community is, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialised RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

Further risks that were later discussed include an intensified replication crisis,
making research less accessible, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
and LLM-assisted data analysis to become less reproducible.
This is not an entirely new issue, and similar concerns exist in the HPC community,
where reproducing simulation data and data analysis workflows presents tough challenges~\cite{Antunes2024}.
Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
When applicable, standardised and version-controlled models should be preferred
over self-developed models, and self-developed models should be treated like any
other research output (i.e. publish the trained model, training data and training workflow).
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more likely to be driven by institutions that are able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}
and the malicious use of hidden prompts~\cite{Gibney2025}.
As researchers increase their reliance on AI assistants,
RSEs will be freed from repeatedly crafting similar individual elements,
but will be spending more of their time designing and reviewing software written by other humans and AI tools.
To address these risks, the RSE of the future should not only steer every part
of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
but also contribute to endangered community resources and infrastructure that
will keep research accessible to the wider scientific community.

@@

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}

@@

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognise research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.
